---
title: "Case report: CEDNIK SYNDROME"
format: docx
editor: R Studio - Quatro
autors: M. Ewen, M. Gezmen, F. Paz
bibliography: wes.bib
---

## Aims

You will receive a set of exome data for a family trio to identify a disease-causing variant using the tools presented in the course. Your task for this case report is to summarize the steps leading to the causal variant and to reflect on how you get there. Your report should be reproducible using RMarkdown and code where possible.

Minimally include two graphs and two computed tables in the Results sections that are purely produced from code. Use computed results in the text at least twice. There are many possibilities in this report.

### Extra efforts, not required

-   Use of a bibliography
-   All figures produced from code
-   All numbers produced from code

### Collaboration

-   You are free to support each other in programming but do not exchange code blocks. There will be questions on the code you wrote. In exceptional cases you can indicate that code was developed by others.
-   If you are stuck, reach out to the tutors.
-   If you collaborate, please indicate in the report.

```{r libraries, include = FALSE}
library(tidyverse)
```

## Introduction

Neurocutaneous syndrome, characterized by cerebral dysgenesis, neuropathy, ichthyosis, and keratoderma, or CEDNIK syndrome for short, is a autosomal recessive genetic disorder (Hsu et al. 2017). This disease is caused by a loss-of-function variant in SNAP29 (synaptosomal-associated protein 29), which encodes a SNARE protein, a vesicular membrane fusion protein (Schiller et al. 2016). Symptoms of the disease Microcephaly, Global developmental delay, Ataxia, Ichthyosis, Diffuse palmoplantar hyperkeratosis.

This syndrome was first discovered in 2005— within the 22q11.2 region and the SNAP29 gene —using homozygosity mapping on two unrelated families with a history of consanguineous marriage that exhibited similar symptoms (Sprecher et al. 2005). In the subsequent years, up to 2021, nearly 20 variants of this syndrome were discovered. (Mah-Som et al. 2021). 

CEDNIK syndrome is an extremely rare disorder, with an estimated prevalence of less than 1 per 1,000,000 individuals; a reliable population-based incidence has not been established due to the small number of reported cases (Orphanet n.d.). 

Variant detection and diagnosis have become much more accessible with the advent of next-generation sequencing (NGS). Unlike earlier sequencing technologies, NGS enables the simultaneous analysis of anything from a few genes to hundreds of genes, the entire exome, or even the whole genome. NGS systems employ approaches such as whole-genome sequencing (WGS) and whole-exome sequencing (WES) (Brancato et al. 2025).

In this report, we focus on WES because Sydnik syndrome is a Mendelian monogenic disorder; identifying variations within the exons alone is generally sufficient for its detection. In the case we analyzed, a 3-year-old girl exhibited symptoms of the syndrome while her parents were healthy; therefore, our objective was to perform WES for this family of three and identify the variant by comparing the results against the reference human genome (GRCh37/hg19). This type of analytical method is known as "trio analysis," (Gabriel et al. 2022) and given that this syndrome is an autosomal recessive disorder, we can infer that both parents are heterozygous carriers (Hsu et al. 2017).


## Materials and methods
```{r}

```

### Study Participants/Whole exome sequencing

-   Description of the Corpas family project (@corpas_crowdsourced_2015)
    -   Also check https://manuelcorpas.com/2016/05/24/my-personal-exome-analysis-part-i-first-findings-2/ )
-   Draw pedigree (mark affected child) and give alleles per individual for the variant that you found. You can draw the pedigree by hand but or use a web programs. Pedigrees can be drawn by programs, including R but the effort required is not worth for a trio.
-   How was the sequencing data generated, on which machine, using which technology, using which sequencing kit, exome array?

### NGS data preprocessing

<<<<<<< Updated upstream
Next-Generation Sequencing (NGS) data preprocessing was used to refine and align raw reads prior to downstream Whole-Exome Sequencing (WES) analysis. The input dataset comprised individual raw FASTQ files corresponding to the family trio: mother, father, and affected child, from the Corpasome family. This tools are organized for each person as .fq files as paired-ends reads, 150 bp per read for R1 and R2.

```{r}
#read <- files
```

For the Quality evaluation, we access to the "FastQC Report (version 0.12.1)" for each file. Using the raw data (.fq data files) as an input, we can graph and see the organization of the sequences, %GC percentage and the quality scores. We considerate the Q scores above 30 (Q30), in acordance to the standart Illumina quality trimming sequence [@illumina].

After the confirmation of the good quality of the sequences, we proceed with the allignment and trimming of information.

We processed the quantification of bp related to the harmful variants. for finding differential gene expresions.

```{r}
#Describe data formats used and give overview of methods with a figure showing the NGS processing workflow (could be from other publications, if so please reference): from raw fastq files to called variants.


```

- Quality control: describe method and parameters to evaluate quality of the raw sequencing data

- NGS processing:

  - Describe mapping method that was used, which reference genome, ..
  - Describe GATK pipeline (see GATK webpage for documentation)
  - Describe the different steps, what are they doing, what is the input/output.
=======
-   Describe data formats used and give overview of methods with a figure showing the NGS processing workflow (could be from other publications, if so please reference): from raw fastq files to called variants.
-   Quality control: describe method and parameters to evaluate quality of the raw sequencing data
-   NGS processing:
    -   Describe mapping method that was used, which reference genome, ..
    -   Describe GATK pipeline (see GATK webpage for documentation)
    -   Describe the different steps, what are they doing, what is the input/output.
>>>>>>> Stashed changes

### Variant annotation and filtering

-   How were variants called?
-   Which parameter were used?
    -   Describe Annovar annotation pipeline. Which annotations and databases were used for annotation?
-   Variant filter strategy:
    -   What annotations/parameters have been used for filtering down the variants?

## Results

### QC

-   Summarize QC results from Fastqc: (1 paragraph)
    -   one table summarizing read statistics for all three samples and
    -   show one read quality metric for all samples in one figure.

### Sample processing

-   Table with read statistics (read numbers for each processing step including total number, after initial mapping, after duplicate removal, etc.). What is the outcome of the NGS workflow?

### Variant calling/filtering

-   How many variants were in the raw sample output and then called in the initial `bcf` files, in the `vcf` file filtered for read depth and finally in the demultiplexed `vcf` file? What means demultiplexing here?
-   How many variants were finally called over all samples and annotated using ANNOVAR? How many variants remaining after each filtering step (1 or 2 tables)? Describe impact of each filtering step by number of qualifying variants.
-   How many exonic/intronic variants were called? (1-2 sentences, 1 table). Were there any variants related to alternative splicing?
-   How many different exonic variant types (nonsynonymous/synonymous/frameshift/non-frameshiftindels/etc) were called? (1-2 sentences, 1 table).
-   QC measures: calculate and discuss
    -   transistion/transversion ratio
    -   nonsynonymous/synonymous ratio
-   Were variants found in candidate variants genes? If yes, with how many reads covered, what was the allele frequency of the alternative allele. Discuss quality of the calls. Report sequencing depth for all samples for candidate variant.
-   Report gene coverage for all samples for candidate gene.
-   Show variant in IGV from all three family members. Which mode of inheritance you have found and what is the support for it?
-   Describe candidate variant and gene using databases.
-   Check the current allele frequencies in gnoMAD for your variants (web)
-   Annotate your variant with more scores and annotation using [wANNOVAR](http://wannovar.wglab.org/) and discuss if they support the hypothesis that the variant is disease causing
-   Collect data about your variant and gene and the position of your variant from GeneCards, Uniprot, OMIM, GTEX. Is the variant located in a certain protein domain, does it is region where other disease-causing variants are located, is the protein/gene expressed in the tissue that you would expect?
-   Use different tools for variant interpretation like MutationMapper, MuPIT, MTR-viewer, VSranker, ClinVar
-   Have at least one figure that shows the location of the variant on the protein sequence, e.g. using MutationMapper or the R package [G3viz](https://cran.rstudio.com/web/packages/g3viz/vignettes/introduction.html)

## Discussion

-   Summary of analysis results.

-   Which variant and gene been found?

-   Is the mode of inheritance of your variant known or not known?

-   What evidence you have found that supports your hypothesis that you have found a causative variant?

-   Can you think of experiments that could show that you have found a disease-causing variant?

## References
