---
title: "Case report: CEDNIK SYNDROME"
format: html
autors: "M. Ewen, M.Gezman, F. Paz"
editor: source
bibliography: wes.bib
csl: cell.csl
echo: false
---

## Aims

You will receive a set of exome data for a family trio to identify a disease-causing variant using the tools presented in the course. Your task for this case report is to summarize the steps leading to the causal variant and to reflect on how you get there. Your report should be reproducible using RMarkdown and code where possible.

Minimally include two graphs and two computed tables in the Results sections that are purely produced from code. Use computed results in the text at least twice. There are many possibilities in this report.

### Extra efforts, not required

- Use of a bibliography
- All figures produced from code
- All numbers produced from code

### Collaboration

- You are free to support each other in programming but do not exchange code blocks. There will be questions on the code you wrote. In exceptional cases you can indicate that code was developed by others.
- If you are stuck, reach out to the tutors.
- If you collaborate, please indicate in the report.

```{r libraries, include = FALSE}
library(tidyverse)
library(ggplot2)
library(gt)
library(quadprog)
library(kinship2)

#for the workflow graphics
#library(igraph)
library(showtext)
library(rcartocolor)
getwd()
#Library from the HPC section/WEScourse/disease/corpasome
#Table for the complete filtered chromosome
chromosomes <- read_tsv("data/corpasome.trio.chr22.snps.filtered.annovar.tsv")
chromosomes

main_folder <- "data/fastqc"
variants <- read_tsv("data/corpasome.trio.chr22.snps.filtered.annovar.tsv", col_types = cols(clinvar_20140702 = "c"))
```

```{r, include = FALSE}
#Chunk 0
#Table for the variants
variants <- read_tsv("data/corpasome.trio.chr22.snps.filtered.annovar.tsv", col_types = cols(clinvar_20140702 = "c"))

#Table for AP variants (joints)
variants_ap <- read_tsv("data/corpasome.trio.chr22.snps.filtered.annovar.tsv", 
  col_types = list("clinvar_20140702" = "c")) |>
  select(Start, End, Ref, Alt,  Gene.refGene, INFO) |> 
  filter(str_detect(Gene.refGene , "^AP")) |>  
  distinct()
variants_ap #just to print it
```

## Introduction

Neurocutaneous syndrome, characterized by cerebral dysgenesis, neuropathy, ichthyosis, and keratoderma, or CEDNIK syndrome for short, is a autosomal recessive genetic disorder [@hsu2017cednik]. This disease is caused by a loss-of-function variant in SNAP29 (synaptosomal-associated protein 29), which encodes a SNARE protein, a vesicular membrane fusion protein [@schiller2016establishment]. Symptoms of the disease Microcephaly, Global developmental delay, Ataxia, Ichthyosis, Diffuse palmoplantar hyperkeratosis.

This syndrome was first discovered in 2005— within the 22q11.2 region and the SNAP29 gene —using homozygosity mapping on two unrelated families with a history of consanguineous marriage that exhibited similar symptoms [@sprecher2005mutation]. In the subsequent years, up to 2021, nearly 20 variants of this syndrome were discovered [@mah2021new].

CEDNIK syndrome is an extremely rare disorder, with an estimated prevalence of less than 1 per 1,000,000 individuals; a reliable population-based incidence has not been established due to the small number of reported cases [@orphanetcednik].

Variant detection and diagnosis have become much more accessible with the advent of next-generation sequencing (NGS). Unlike earlier sequencing technologies, NGS enables the simultaneous analysis of anything from a few genes to hundreds of genes, the entire exome, or even the whole genome. NGS systems employ approaches such as whole-genome sequencing (WGS) and whole-exome sequencing (WES) [@brancato2025ngs].

In this report, we focus on WES because Sydnik syndrome is a Mendelian monogenic disorder; identifying variations within the exons alone is generally sufficient for its detection. In the case we analyzed, a 3-year-old girl exhibited symptoms of the syndrome while her parents were healthy; therefore, our objective was to perform WES for this family of three and identify the variant by comparing the results against the reference human genome (GRCh37/hg19). This type of analytical method is known as "trio analysis" [@gabriel2022trio], and given that this syndrome is an autosomal recessive disorder, we can infer that both parents are heterozygous carriers [@hsu2017cednik].

### Study Participants/Whole exome sequencing

```{r}
#Description of the Corpas family project [@corpas_crowdsourced_2015]
  #Also check https://manuelcorpas.com/2016/05/24/my-personal-exome-analysis-part-i-first-findings-2/ )
  #Draw pedigree (mark affected child) and give alleles per individual for the variant that you found. You can draw the pedigree by hand but or use a web programs. Pedigrees can be drawn by programs, including R but the effort required is not worth for a trio.
  #How was the sequencing data generated, on which machine, using which technology, using which sequencing kit, exome array?
```

The base of our work is publicly available genetic data from a Spanish family, here consisting of a father, mother and their daughter, referred to as "the Corpasome" [@corpas_crowdsourced_2015]. In our case the genome of the daughter is synthetically altered and a disease phenotype called CEDNIK Syndrome, caused by these alterations, is diagnosed. Main symptoms are developmental delay and skin disorder. The latter mainfasting at the age of two in ichthyosis and thickening of palms and soles. Further, she has been diagnosed with microcephaly and poor motor development as well as impaired coordination. As Cednik Syndrome is an autosomal recessive inheritable disease and the parents are healthy, the father and mother both have one chromosome with the disease-causing variant and one without. After deciding to make their genomes publicly available, the Corpas family bought direct-to-consumer sequencing privately to collect DNA in saliva using Oragene kits. Exomes were captured with Agilent SureSelect Human All Exon (44 Mb design) and sequenced on Illumina HiSeq with an average coverage of about 30x over the target region.

```{r}
#| warning: false
ped <- read_table("data/corpasome.ped",
                  col_names = c("family", "id", "father", "mother", "sex", "phenotype"),
                  col_types = "ccccii")
ped2 <- ped |>
  mutate(across(c(id, father, mother), \(x) str_remove(x, "^corpasome_")),
         father   = na_if(father, "0"),     # 0 = no parent in the file -> NA
         mother   = na_if(mother, "0"),
         affected = as.integer(phenotype == 2))  # 2 = affected -> 1, otherwise 0

pedi <- pedigree(id      = ped2$id,
                 dadid   = ped2$father,
                 momid   = ped2$mother,
                 sex     = ped2$sex,          # 1 = male, 2 = female
                 affected = ped2$affected)

plot(pedi)
```

### NGS data preprocessing

```{r}
#Describe data formats used and give overview of methods with a figure showing the NGS processing workflow (could be from other publications, if so please reference): from raw fastq files to called variants.

#Quality control: describe method and parameters to evaluate quality of the raw sequencing data

#NGS processing:

  #- Describe mapping method that was used, which reference genome, ..
  #- Describe GATK pipeline (see GATK webpage for documentation)
  #- Describe the different steps, what are they doing, what is the input/output.

```

Next-Generation Sequencing (NGS) data preprocessing was used to refine and align raw reads prior to downstream Whole-Exome Sequencing (WES) analysis. The input dataset comprised individual raw FASTQ files corresponding to the family trio: mother, father, and affected child, from the Corpasome family. This tools are organized for each person as .fq files as paired-ends reads, 150 bp per read for R1 and R2.

For the Quality evaluation, we access to the "FastQC Report (version 0.12.1)" for each file. Using the raw data (.fq data files) as an input, we can graph and see the organization of the sequences, %GC percentage and the quality scores. We considerate the Q scores above 30 (Q30), in acordance to the standart Illumina quality trimming sequence [@illumina].

After the confirmation of the good quality of the sequences, we proceed with the allignment and trimming of information.

We processed the quantification of bp related to the harmful variants. for finding differential gene expresions.

### Variant annotation and filtering

```{r}
#How were variants called?

#Which parameter were used?
  
  #Describe Annovar annotation pipeline. Which annotations and databases were used for annotation?
  
  #Variant filter strategy:
  
  #What annotations/parameters have been used for filtering down the variants?
```

## Results

### QC

```{r}
#| label: "fastqc"

# Example folder, change this 
fastqc_base_dir <- "data/fastqc/"

# the directory fastqc contains other data such as zip and html files
# 
valid_dirs <- str_subset(dir(fastqc_base_dir), "fastqc$") #gets directories for relevant folders (with fastqc at the end)

processed <- read_tsv(str_c(fastqc_base_dir, 
                            valid_dirs, 
                            "/fastqc_data.txt"), #str_c pastes base dir, each folder in valid dir and txt together --> real all of them and stack together in one table
        col_names = TRUE, #use third line as columnnames
        skip = 2, #skips first 2 rows of each file
        n_max = 8,#read only 8 next lines (basic statistics)
        id = "file_name_long", #adds column with source file path for each row
        show_col_types = FALSE) |> #suppresses message about coltypes
#new column before source path, regex extracts file name  
  mutate(file_name = str_extract(file_name_long, "/([^/]+?)_", group = 1), .before = file_name_long) |> 
  select(!file_name_long) |> #drop this column, file name can be used
  #long to wide: id_cols = file_name gives one row per sample, table is now filled with characters
  pivot_wider(id_cols = file_name, names_from =`#Measure` , values_from = Value) |> 
  select(Filename, `Total Sequences`, `Total Bases`, `Sequence length`, `%GC`) #keep only wanted samples

gt(processed) #gt formats a tibble as a presentation table for report


```

To begin the WES analysis, it is important to build on raw-read data that fulfills the quality requirements. By assessing the quality of the fastq file, it can be assured that later interpretation is not based on technical errors. Regarding the Corpasome data, the sequence length for all samples is as expected as well as the GC percentage with small variations in the reads of the mother displaying a percentage of 44 compared to 45% in the father and daughter. The number of total sequences as well as the number of bases measured is the same for read 1 and read 2 in each family member, whereas the number varies between different persons. The maximal difference can be found between the father and the daughter where the father has 6.3 Mbp more. These differences, however, lie within the natural range of variation of genome sizes. (check reference!)

```{r qc_boxplots}
#genereate needed qc file with all pb_qual files and their source
files <- list.files(main_folder, pattern = "fastqc_data.txt",
                    recursive = TRUE, full.names = TRUE)

qc <- files |>
  set_names() |>
  map(\(f) read_tsv(f, skip = 13, n_max = 50, show_col_types = FALSE)) |>
  list_rbind(names_to = "file") |>
  mutate(folder      = basename(dirname(file)),
         read_set    = str_c("Read ", str_extract(folder, "(\\d+)_fastqc$", group = 1)),   
         sample      = str_remove(folder, "\\.\\d+_fastqc$"),
         base_factor = factor(`#Base`, levels = unique(`#Base`))) |>
  select(!file)

qc |>
  group_by(folder) |> #mutate runs separately for each folder
  mutate(`#Base` = factor(`#Base`, levels = `#Base`)) |> #turn position in a factor to avoid alphabetical sorting
  
  #generate boxplot
  ggplot(aes(x = base_factor, y = Mean, fill = read_set)) +
  geom_boxplot(
    aes(
      ymin = `10th Percentile`,
      lower = `Lower Quartile`,
      middle = `Median`,
      upper = `Upper Quartile`,
      ymax = `90th Percentile`), stat = "identity", staplewidth = 0.5) +
  
  #add threshold lines
   geom_hline(aes(yintercept = 28),
              linetype = "dashed",
              color = "orange") + # Good quality threshold
  geom_hline(aes(yintercept = 20),
             linetype = "dashed",
             color = "red") +
   geom_line(aes(x = base_factor, y = Mean, group = read_set),
              linetype = "solid",
              color = "blue") +
  labs(title = "Quality scores across all bases", x = "Position in read (bp)", y = "Mean Quality Score") +
  theme_minimal() +
  facet_wrap(vars(folder), ncol = 2) +  #makes one panel per read set
  # scale_x_discrete(breaks = qc$`#Base`[seq(1, nrow(qc), by = 10)]) +
  scale_fill_discrete(name = "Folder") + #legend hidden anyways?
  scale_x_discrete(breaks = seq(1, 100, 5)) + #show every 5th label
  coord_cartesian(ylim = c(0, 45)) +
  theme(legend.position = "none")

#qc_graph_all
```

The boxplot depicts the mean quality score per position in each read separately for read 1 and read 2 in the father, mother and daughter. For each sample the last positions of some reads fall beneath the threshold of a quality score of 30 while the best quality scores are achieved for positions in the middle of the read. Especially in the reads coming from the mother, quality scores beneath 20 are achieved. However, for all positions in all samples the median remains at a quality score above 30 meaning that no reads need to be shortened for the downstream analysis. Nevertheless, it is important to remark that the samples derived from the mother show the lowest quality across different metrics.(which ones)

From the quality control regarding the raw sequencing data, there is no detectable issue that would affect further analysis. Furthermore, in the following steps, more metrics will be assessed to assure correct variant calling.

### Sample processing

```{r}
bam_stats <- read_delim("data/bam_statistics.txt", col_names = c("nreads", "file"),  delim = " ", show_col_types = FALSE)

step_labels <- c(
  "bam"         = "Aligned and sorted",
  "markdup.bam" = "Duplicates removal",
  "realn.bam"   = "Indel realignment",
  "recal.bam"   = "Base quality recalibration"
)

bam_base <- bam_stats |>
  mutate(person   = str_extract(file, "^\\w+"),
         bam_step = str_extract(file, "\\D+$"),
         bam_step = str_remove(bam_step, "^\\.")) |>
  pivot_wider(id_cols = -file, names_from = person, values_from = nreads) |>
  rename(Filename = bam_step) |>
  mutate(description = unname(step_labels[Filename]), .after = Filename)

removed_row <- bam_base |>
  filter(Filename %in% c("bam", "recal.bam")) |>
  summarise(across(c(dad, mom, daughter),
                   \(x) x[Filename == "bam"] - x[Filename == "recal.bam"])) |>
  mutate(Filename = "",
         description = "Removed (total)",
         .before = dad)

pct_row <- bam_base |>
  filter(Filename %in% c("bam", "recal.bam")) |>
  summarise(across(c(dad, mom, daughter),
                   \(x) round(100 * (x[Filename == "bam"] - x[Filename == "recal.bam"]) /
                                x[Filename == "bam"], 1))) |>
  mutate(Filename = "",
         description = "Removed (%)",
         .before = dad)


bam_table <- bind_rows(bam_base, removed_row, pct_row) 

gt(bam_table)

#add summary rows below
removed <- bam_table |> filter(description == "Removed (total)")
pct     <- bam_table |> filter(description == "Removed (%)")

pct_range <- range(unlist(pct[, c("dad", "mom", "daughter")]))

#to format numbers in text 
fmt <- \(x) format(x, scientific = FALSE, big.mark = ",", trim = TRUE)
```

The table displays the different steps of sample processing and the effect on the number of alignments. As expected, only the duplication removal reduces the number of records. In total `r fmt(removed$dad)` alignments have been removed in the sample of the dad, `r fmt(removed$mom)` for the mom and `r fmt(removed$daughter)` for the daughter, which is in a range of `r pct_range[1]`% to `r pct_range[2]`% of all alignments in the unprocessed file.

### Variant calling/filtering

To begin the variant calling, the raw file includes all covered positions in exons which could potentially contain variants. It includes all three samples and displays 1,418,395 positions.

- How many variants were in the raw sample output and then called in the initial `bcf` files, in the `vcf` file filtered for read depth and finally in the demultiplexed `vcf` file? What means demultiplexing here?
- How many variants were finally called over all samples and annotated using ANNOVAR? How many variants remaining after each filtering step (1 or 2 tables)? Describe impact of each filtering step by number of qualifying variants.
- How many exonic/intronic variants were called? (1-2 sentences, 1 table). Were there any variants related to alternative splicing?
- How many different exonic variant types (nonsynonymous/synonymous/frameshift/non-frameshiftindels/etc) were called? (1-2 sentences, 1 table).
- QC measures: calculate and discuss
  - transistion/transversion ratio
  - nonsynonymous/synonymous ratio
- Were variants found in candidate variants genes? If yes, with how many reads covered, what was the allele frequency of the alternative allele. Discuss quality of the calls. Report sequencing depth for all samples for candidate variant.
- Report gene coverage for all samples for candidate gene.
- Show variant in IGV from all three family members. Which mode of inheritance you have found and what is the support for it?
- Describe candidate variant and gene using databases.
- Check the current allele frequencies in gnoMAD for your variants (web)
- Annotate your variant with more scores and annotation using [wANNOVAR](http://wannovar.wglab.org/) and discuss if they support the hypothesis that the variant is disease causing
- Collect data about your variant and gene and the position of your variant from GeneCards, Uniprot, OMIM, GTEX. Is the variant located in a certain protein domain, does it is region where other disease-causing variants are located, is the protein/gene expressed in the tissue that you would expect?
- Use different tools for variant interpretation like MutationMapper, MuPIT, MTR-viewer, VSranker, ClinVar
- Have at least one figure that shows the location of the variant on the protein sequence, e.g. using MutationMapper or the R package [G3viz](https://cran.rstudio.com/web/packages/g3viz/vignettes/introduction.html)

## Discussion

- Summary of analysis results.

- Which variant and gene been found?

- Is the mode of inheritance of your variant known or not known?

- What evidence you have found that supports your hypothesis that you have found a causative variant?

- Can you think of experiments that could show that you have found a disease-causing variant?

## References
